---
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    highlight: zenburn
    theme: flatly
---

---
title: "Differential expression"
author: ""
date: "`r Sys.Date()`"
bibliography: bcbioRnaseq.bib
---

```{r setup, message=FALSE}
library(DESeq2)
library(bcbioRnaseq)

downloads()

source("setup.R") # modify parameters before exec this line
# Contrast
# 1. Design matrix parameter.
# 2. Numerator for LFC (expt).
# 3. Denominator for LFC (control).
# @seealso [DESeq2::results()]
contrast <- c("gender", "male", "female")

# Design formula
design <- formula(~ gender+genotype+treatment) # change to desire column

# Significance cutoffs
alpha <- 0.05
lfc <- 1
```

```{r header, child="header.Rmd"}
load("data/bcb.rda")
```


```{r dds-de, results="hide", eval=sum(ls() == "ddsde")==0}
txi <- bcbio(bcb, "tximport")
coldata <- colData(bcb)
coldata$treatment = relevel(coldata$treatment, ref="untreated")
coldata$genotype = relevel(coldata$genotype, ref="wt")
coldata$gender = relevel(coldata$gender, ref="male")

ddsde <- DESeqDataSetFromTximport(
  txi = txi,
  colData = coldata,
  design = design)
ddsde <- DESeq(ddsde)
rld <- rlog(ddsde)
save_data(ddsde, dir=data_out)
resultsNames(ddsde)
sizeFactors(ddsde)
```

# Contrasts to look at

```
All of the original experiments contained male hepatocytes under the following
treatments:

1) Normal - no treatment
2) Normal - CCl4
3) yap/taz knockout
4) Yap/taz knockout - CCl4

If we can compare these male and female (normal) hepatocytes with condition 1) that would be great. I’m hoping that if we did a PCA plot, with the prior and current mice, then the males would group together and females would group elsewhere. A PCA plot and a gene heat map would be more than enough for my grant.

At worst, if the batch effect is too difficult to overcome in a short time, we could just compare the expression pattern from the 3 males and 5 females that will be sequenced on the same lane in this experiment.

I am envisioning that this data would be a figure before the Yap/Taz manipulations in the current paper as well, showing the differences that sex may play.
```

Dean wants to look at gender specific differences in hepatocytes, here is what
we have for samples in terms of genotype, gender and treatment.

```{r sample-count-plot, results='asis'}
library(dplyr)
sample_counts = colData(bcb) %>%
  as.data.frame() %>%
  group_by(gender, genotype, treatment) %>%
  summarise(count=n())
knitr::kable(sample_counts)
```

# Clustering
The samples separate by gender and we can see the batches of untreated male
samples cluster together, so the gender-specific differences are stronger than
any batch effect that might be occuring. That is a good sign for trying to find
differences between the female and male samples.

```{r clustering}
vst = varianceStabilizingTransformation(ddsde)
pca_loadings = function(object, ntop=500) {
  rv <- matrixStats::rowVars(assay(object))
  select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,
      length(rv)))]
  pca <- prcomp(t(assay(object)[select, ]))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  names(percentVar) = colnames(pca$x)
  pca$percentVar = percentVar
  return(pca)}
pc = pca_loadings(vst)
comps = data.frame(pc$x)
comps$description = rownames(comps)
coldata$description = rownames(coldata)
library(dplyr)
comps = comps %>% left_join(as.data.frame(coldata), by="description")
comps$batch = ifelse(grepl("_S", comps$description), "second", "first")
comps$gender_batch = paste(comps$gender, comps$batch, sep="_")
colorby = "gender"
pca_plot = function(comps, nc1, nc2, colorby) {
   c1str = paste0("PC", nc1)
   c2str = paste0("PC", nc2)
  if(!(c1str %in% colnames(comps) && c2str %in% colnames(comps))) {
    warning("Higher order components not found, skipping plotting.")
    return(NA)
  }
  ggplot(comps, aes_string(c1str, c2str, color=colorby)) +
    geom_point() + theme_bw() +
    facet_wrap(~genotype+treatment) +
    xlab(paste0(c1str, ": ", round(pc$percentVar[nc1] * 100), "% variance")) +
    ylab(paste0(c2str, ": ", round(pc$percentVar[nc2] * 100), "% variance"))
}
pca_plot(comps, 1, 2, "gender")
```

# Effect of gender
We fit a model that takes into account the effect of genotype, the effect of
treatment and the effect of gender and look at the effect of gender, controlling
for all of the other effects.

# Alpha level (FDR) cutoffs

Let's take a look at the number of genes we get with different false discovery rate (FDR) cutoffs. These tests subset *P* values that have been multiple test corrected using the Benjamini Hochberg (BH) method [@Benjamini:1995ws].

```{r alpha_summary}
contrast <- c("gender", "female", "male")
gender <- results(ddsde, contrast=c("gender", "female", "male")) %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var="ensgene") %>%
  left_join(as.data.frame(rowData(bcb)), by="ensgene")
gendersig <- subset(gender, padj < 0.05)
gendersighigh <- subset(gender, padj < 0.05 & abs(log2FoldChange) > 1)
```

There are `r nrow(gendersig)` genes that vary with gender at FDR < 0.05,
and there are `r nrow(gendersighigh)` that have a fold change greater than
two-fold.

As a sanity check, we should be seeing Xist as higher in the female. We can
see that is true below (positive log2FoldChange is higher in the females for
this comparison).

```{r check-xist}
subset(gender, symbol == "Xist")
```

# Plots

## Mean average (MA)

An MA plot compares transformed counts on `M` (log ratio) and `A` (mean average) scales [@Yang:2002ty].

```{r plot_ma}
label = gender %>% arrange(padj) %>% head(25)
plot_ma(gender, label_points=label$symbol) + theme_light(base_size=10)
```


## Volcano

A volcano plot compares significance (BH-adjusted *P* value) against fold change (log2) [@Cui:2003kh; @Li:2014fv]. Genes in the green box with text labels have an adjusted *P* value are likely to be the top candidate genes of interest.

```{r plot_volcano}
res = results(ddsde, contrast=c("gender", "female", "male"))
plot_volcano(bcb, res = res, lfc = lfc)
```

## Heatmap

This plot shows only differentially expressed genes on a per-sample basis. We have scaled the data by row and used the `ward.D2` method for clustering [@WardJr:1963eu].

```{r plot_deg_heatmap}
plot_deg_heatmap(bcb, res = res, dt = rld, lfc = lfc)
```

Below is a heatmap of just the untreated, wildtype samples, so just looking at
the affect of gender.

```{r heatmap-untreated-only}
untreated = bcb[, colData(bcb)$treatment == "untreated" & colData(bcb)$genotype == "wt"]
untreateddds = ddsde[, colData(bcb)$treatment == "untreated" & colData(bcb)$genotype == "wt"]
rld = rlog(untreateddds)
plot_deg_heatmap(untreated, res=res, dt=rld, lfc=lfc)
```

```{r write-counts}
tpm = txi$abundance %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensgene")
counts = txi$counts %>%
  as.data.frame() %>%
  tibble::rownames_to_column("ensgene")
readr::write_csv(tpm, "results/tpm.csv.gz")
readr::write_csv(counts, "results/counts.csv.gz")
```

## GSEA
```{r gsea-analysis}
orgdb = "org.Mm.eg.db"
library(dplyr)
library(clusterProfiler)
library(orgdb, character.only=TRUE)
library(biomaRt)
biomart_dataset = "mmusculus_gene_ensembl"
keggname = "mmu"
mart = biomaRt::useMart(biomart = "ensembl", dataset=biomart_dataset)
symbols = biomaRt::getBM(attributes=c("ensembl_gene_id", "mgi_symbol"), mart=mart)
entrez = biomaRt::getBM(attributes = c("ensembl_gene_id", "entrezgene"), mart=mart)
entrezsymbol = biomaRt::getBM(attributes=c("mgi_symbol", "entrezgene"), mart=mart)
entrezsymbol$entrezgene = as.character(entrezsymbol$entrezgene)
entrez$entrezgene = as.character(entrez$entrezgene)

summarize_cp = function(res, comparison) {
  summaries = data.frame()
  for(ont in names(res)) {
     ontsum = summary(res[[ont]])
     ontsum$ont = ont
     summaries = rbind(summaries, ontsum)
  }
  summaries$comparison = comparison
  return(summaries)
}

gsea_cp = function(res, comparison) {
  res = res %>%
    as.data.frame() %>%
    tibble::rownames_to_column("ensembl_gene_id") %>%
    left_join(entrez, by="ensembl_gene_id") %>%
    filter(!is.na(entrezgene)) %>%
    filter(!is.na(log2FoldChange)) %>%
    filter(!is.na(lfcSE))
  fccol = ifelse("lfcMLE" %in% colnames(res), "lfcMLE", "log2FoldChange")
  lfc = data.frame(res)[, fccol]
  lfcse = data.frame(res)[, "lfcSE"]
  genes = lfc/lfcse
  names(genes) = res$entrezgene
  genes = genes[order(genes, decreasing=TRUE)]
  cc = gseGO(genes, ont="CC", OrgDb=orgdb,
             nPerm=500, pvalueCutoff=1, pAdjustMethod="BH",
             verbose=TRUE)
  mf = gseGO(genes, ont="MF", OrgDb=orgdb,
             nPerm=500, pvalueCutoff=1, pAdjustMethod="BH",
             verbose=TRUE)
  bp = gseGO(genes, ont="bp", OrgDb=orgdb,
             nPerm=500, pvalueCutoff=1, pAdjustMethod="BH",
             verbose=TRUE)
  genes = data.frame(res)[, fccol]
  names(genes) = res$entrezgene
  genes = genes[order(genes, decreasing=TRUE)]
  genes = genes[!is.na(genes)]
  kg = gseKEGG(geneList=genes, organism=keggname, nPerm=500,
               pvalueCutoff=1, verbose=TRUE)
  if(orgdb == "org.Hs.eg.db") {
    do = summary(gseDO(geneList=genes, nPerm=500, pvalueCutoff=1,
                       pAdjustMethod="BH", verbose=TRUE))
    do$ont = "DO"
    all = list(mf=mf, cc=cc, bp=bp, kg=kg, do=do)
  }
  else {
    all = list(mf=mf, cc=cc, bp=bp, kg=kg)
  }
  all[["summary"]] = summarize_cp(all, comparison)
  return(all)}

convert_core_ids = function(res) {
  res = res %>%
    mutate(geneID=strsplit(as.character(geneID), "/")) %>%
    tidyr::unnest(geneID) %>%
    left_join(entrezsymbol, by=c("geneID"="entrezgene")) %>%
    group_by(ID, Description, GeneRatio, BgRatio, pvalue, p.adjust, qvalue,
            Count, ont, comparison) %>%
    summarise(geneID=paste(geneID, collapse="/"),
              symbol=paste(mgi_symbol, collapse="/"))
  return(res)
  }
direction_breakdown = function(res, deseq2res) {
  ocols = colnames(res)
  deseq2res = deseq2res %>%
    as.data.frame() %>%
    tibble::rownames_to_column() %>%
    dplyr::select(log2FoldChange, rowname)
  ids = res %>% mutate(symbol=strsplit(as.character(symbol), "/")) %>%
    tidyr::unnest(symbol) %>%
    left_join(symbols, by=c("symbol"="mgi_symbol")) %>%
    left_join(deseq2res, by=c("ensembl_gene_id"="rowname")) %>%
    mutate(direction=ifelse(log2FoldChange > 0, "up", "down"))
  ids = ids[, c(ocols, "direction")]
  ids = ids %>% group_by(ID, Description, GeneRatio, BgRatio, pvalue,
                         p.adjust, qvalue, Count, ont, pvalue, p.adjust,
                         qvalue, comparison, direction) %>%
    summarise(symbols=paste(symbol, collapse="/"))
  return(ids)}
signal_to_noise = function(res) {
  res = res %>%
    as.data.frame() %>%
    tibble::rownames_to_column("ensembl_gene_id") %>%
    left_join(entrez, by="ensembl_gene_id") %>%
    filter(!is.na(entrezgene)) %>%
    filter(!is.na(log2FoldChange)) %>%
    filter(!is.na(lfcSE))
  fccol = ifelse("lfcMLE" %in% colnames(res), "lfcMLE", "log2FoldChange")
  lfc = data.frame
  lfcse = data.frame(res)[, "lfcSE"]
  genes = lfc/lfcse
  names(genes) = res$entrezgene
  genes = genes[order(genes, decreasing=TRUE)]
  return(genes)
}
convert_gsea_ids = function(res, entrezsymbol) {
 res %>%
  mutate(geneID=strsplit(as.character(core_enrichment), "/")) %>%
    tidyr::unnest(geneID) %>%
    left_join(entrezsymbol, by=c("geneID"="entrezgene")) %>%
  group_by(ID, Description, setSize, enrichmentScore, NES, pvalue, p.adjust,
           qvalues, rank, leading_edge, core_enrichment, ont, comparison) %>%
  summarise(geneID=paste(geneID, collapse="/"),
            symbol=paste(mgi_symbol, collapse="/"))
 }

gender = data.frame(gender)
rownames(gender) = gender$ensgene
gender_gsea = gsea_cp(gender, "gender-effect")
gender_gsea_res = convert_gsea_ids(gender_gsea$summary, entrezsymbol)
dir.create("results/gsea")
readr::write_csv(gender_gsea_res, "results/gsea/gender-gsea-results.csv")
```

# File downloads

The results are saved as gzip-compressed comma separated values (CSV). Gzip compression is natively supported on [macOS][] and Linux-based operating systems. If you're running Windows, we recommend installing [7-Zip][]. CSV files can be opened in [Excel][] or [RStudio][].


## Count matrices

- [`tpm.csv.gz`](results/tpm.csv.gz):
    Transcripts per million, scaled by effective transcript length and library size,
    suitable for plotting.
- [`counts.csv.gz`](results/counts.csv.gz):
    Only use to perform a new differential expression analysis. These counts
    will vary across samples due to differences in sequencing depth, and have
    not been normalized. Do not use this file for plotting genes.

## Differentially expressed genes (DEG)

- [res_gender_female_vs_male_all_genes.csv.gz](results/de/res_gender_female_vs_male_all_genes.csv.gz): differential expression results for all genes
- [res_gender_female_vs_male_deg.csv.gz](results/de/res_gender_female_vs_male_deg.csv.gz): all genes with at least a 2-fold change and a FDR of 0.05
- [res_gender_female_vs_male_deg_lfc_down.csv.gz](results/de/res_gender_female_vs_male_deg_lfc_down.csv.gz): just genes that are downregulated 2-fold and have a FDR of 0.05
- [res_gender_female_vs_male_deg_lfc_up.csv.gz](results/de/res_gender_female_vs_male_deg_lfc_up.csv.gz): just genes that are upregulated 2-fold and have a FDR of 0.05

For all tables, a positive log2FoldChange is higher in the female samples than
the male samples.

DEG tables are sorted by BH-adjusted P value, and contain the following columns:

- `ensembl_gene_id`: [Ensembl][] gene identifier.
- `base_mean`: Mean of the normalized counts per gene for all samples.
- `log2_fold_change`: log2 fold change.
- `lfc_se`: log2 standard error.
- `stat`: Wald statistic.
- `pvalue`: Walt test *P* value.
- `padj`: BH adjusted Wald test *P* value (corrected for multiple comparisons; FDR).
- `external_gene_name`: [Ensembl][] name (a.k.a. symbol).
- `description`: [Ensembl][] description.
- `gene_biotype`: [Ensembl][] biotype (e.g. `protein_coding`).

## GSEA
- [gender-gsea-results.csv](results/gsea/gender-gsea-results.csv): GSEA results for the gender differences.


```{r res_tbl}
res_tbl <- res_tables(bcb, res = res, lfc = lfc, dir=res_out)
save_data(res_tbl, dir=data_out)
```

```{r footer, child="footer.Rmd"}
```
